Covariate analysis of ROSMAP RNASeq with Quantile and SVA normalisation

### 2016-09-26 code/Rmd/ROSMAP_RNASeq_Qntl_SVA.Rmd
Minor changes (included gene based filtering)

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  mutate(Sampleid = paste(Sampleid, Batch, sep = '_')) %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!(Sampleid %in% c('380_120503_1'))) %>%
  as.data.frame()
```
Sample id 380\_120503\_1 is considered as outlier and removed
```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))

@@

@@

Coexpression of genes 
```{r coexp1, cache=FALSE}
cr = cor(t(EXPR))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Outliers detection
```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(EXPR, scale.=T)

@@

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
```
```{r decompse.normalise.data2, fig.height=8, fig.width=16, results='asis'}
tree = hclust(as.dist(t(EXPR)))
plotDendroAndColors(tree, 
                    colors = labels2colors(COVARIATES.tmp), 
                    dendroLabels = FALSE, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES.tmp))
dev.off()
```

### Significant Covariates
Correlation between pca of normalised mRNA expression and covariates
```{r preAdjusted.covariates, cache=TRUE, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
DM = getDesignMatrix(COVARIATES, Intercept = F)
DESIGN = getDesignMat2Fact(DM$design,FactorCovariates)
rownames(DESIGN) = rownames(COVARIATES)
preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR, DESIGN,'non-normalized', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = designMatVarsToCovars(getCovariatesMap(DM),preAdjustedSigCovars$significantCovars)
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising FPKM with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  rownames(DESIGN) <- rownames(COVARIATES)
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION,
                                                DESIGN,
                                                'all adjusted design',
                                                isKeyPlot=TRUE)

@@

## Calculate the number of surrogate variables
# Get number of surrogate variables (for details of the algorithm refer http://biorxiv.org/content/biorxiv/suppl/2016/05/09/052209.DC1/052209-1.pdf)
# SINGULAR.VAL.DECOM <- svd(RESIDUAL.GENE_EXPRESSION)
# NUM.DEG.FREEDOM = dim(RESIDUAL.GENE_EXPRESSION)[2] - dim(MODEL1)[2] + 1
# SING.VALS <- SINGULAR.VAL.DECOM$d^2/sum(SINGULAR.VAL.DECOM$d)
# SING.VALS = SING.VALS[1:NUM.DEG.FREEDOM]

# # Permute and calculate the permuted singular values
# PERM.SING.VALS <- foreach (i = 1:100,
#          .combine = rbind,
#          .packages = c('limma','CovariateAnalysis'),
#          .export = c('PROCESSED_COUNTS','MODEL1')) %dopar% {
#   PERM.COUNTS <- t(apply(PROCESSED_COUNTS$filteredExprMatrix, 1, sample, replace = FALSE))
#   # Estimate voom weights
#   VOOM.ADJUSTED.PERM.GENE_EXPRESSION = voom(PERM.COUNTS, design=MODEL1, plot=F)
#   
#   # Fit linear model using new weights and new design
#   VOOM.ADJUSTED.PERM.FIT = lmFit(VOOM.ADJUSTED.PERM.GENE_EXPRESSION)
#   
#   # Residuals after normalisation
#   RESIDUAL.PERM.GENE_EXPRESSION = residuals.MArrayLM(VOOM.ADJUSTED.PERM.FIT,VOOM.ADJUSTED.PERM.GENE_EXPRESSION$E)
#   
#   PERM.SINGULAR.VAL.DECOM <- svd(RESIDUAL.PERM.GENE_EXPRESSION)
#   PERM.SING.VALS <- PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2/sum(PERM.SINGULAR.VAL.DECOM$d[1:NUM.DEG.FREEDOM]^2)
# }
# num.sv1 = sum(SING.VALS > t(rowMeans(t(PERM.SING.VALS)))[1,1])

num.sv2 = num.sv(EXPR, MODEL1, method = 'be', seed = 123456)
NUM.SV = num.sv2 #min(num.sv1, num.sv2)

## Compute surrogate variables 
SURR.VAR = sva(EXPR, MODEL1, MODEL0, method = "irw", n.sv = NUM.SV, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

Coexpression of genes 
```{r coexp2, cache=FALSE}
cr = cor(t(RESIDUAL.GENE_EXPRESSION))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION, scale.=T)

>>>>

### Data preprocessing
```{r preprocess.data,cache=TRUE, echo=TRUE}
# Remove samples with no cogdx, RIN and PMI scores
METADATA <- METADATA %>%
  ungroup %>%
  mutate(Sampleid = paste(Sampleid, Batch, sep = '_')) %>%
  filter(Sampleid %in% colnames(COUNT)) %>%
  filter(!is.na(cogdx)) %>%
  filter(!is.na(RINcontinuous)) %>%
  filter(!is.na(pmi)) %>%
  filter(!(Sampleid %in% c('380_120503_1', '792_130530_7', '15_120410_3', '256_120426_5'))) %>%
  as.data.frame()
```
Sample id 380\_120503\_1 is considered as outlier and removed
```{r preprocess.data1,cache=TRUE}
indToRetain = intersect(METADATA$Sampleid, colnames(COUNT))

@@

# Set gene counts in specific samples that are deviating 3 sd from other samples to NA
log.mat = apply(EXPR, 1, function(x){
  mn = mean(x, na.rm = T)
  std.dev = sd(x, na.rm = T)
  return((x < (mn-3*std.dev)) | (x > (mn+3*std.dev)))
}) %>% t
EXPR[log.mat] = NA

@@

Coexpression of genes 
```{r coexp1, cache=FALSE}
EXPR.tmp = EXPR
EXPR.tmp[is.na(EXPR)] = 0
cr = cor(t(EXPR.tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Outliers detection
```{r decompse.normalise.data1, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(EXPR.tmp, scale.=T, center = T)

@@

COVARIATES.tmp = data.matrix(COVARIATES)
COVARIATES.tmp[is.na(COVARIATES.tmp)] = 0
```
```{r decompse.normalise.data2, fig.height=8, fig.width=16, results='asis'}
tree = hclust(as.dist(t(EXPR.tmp)))
plotDendroAndColors(tree, 
                    colors = labels2colors(COVARIATES.tmp), 
                    dendroLabels = FALSE, 
                    abHeight = 0.80, 
                    main = "Sample dendrogram",
                    groupLabels = colnames(COVARIATES.tmp))
dev.off()
gc()
```

### Significant Covariates
Correlation between pca of normalised mRNA expression and covariates
```{r preAdjusted.covariates, cache=TRUE, results='asis', echo=FALSE}
# Find correlation between PC's of gene expression with covariates
preAdjustedSigCovars = runPCAandPlotCorrelations(EXPR.tmp, COVARIATES,'non-normalized', isKeyPlot=TRUE)

# Find significant covariates
adjustCovars = preAdjustedSigCovars$significantCovars
```
Significant covariates to adjust at FDR 0.1 are `r adjustCovars`
```{r preAdjustedSigCovars.NULL.ALL, fig.width=15, fig.height=8}
preAdjustedSigCovars[["PC_res"]][[2]]$plotData
```
### Normalisation (iterative design)
Since many covariates are correlated, re-normalising FPKM with an iterative design matrix. Here "cogdx" is chosen as the primary variable of interest
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'cogdx'

@@

  # Find PC of residual gene expression and significant covariates that are highly correlated with PCs
  RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
  RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
  residualSigCovars = runPCAandPlotCorrelations(RESIDUAL.GENE_EXPRESSION.tmp,
                                                COVARIATES,
                                                'all adjusted design',
                                                isKeyPlot=TRUE)

@@

num.sv = num.sv(EXPR.tmp, MODEL1, method = 'be', seed = 123456)

## Compute surrogate variables 
SURR.VAR = sva(EXPR.tmp, MODEL1, MODEL0, method = "irw", n.sv = num.sv, B = 30)$sv
colnames(SURR.VAR) = paste0('SV',1:dim(SURR.VAR)[2])

@@

Coexpression of genes 
```{r coexp2, cache=FALSE}
RESIDUAL.GENE_EXPRESSION.tmp = RESIDUAL.GENE_EXPRESSION
RESIDUAL.GENE_EXPRESSION.tmp[is.na(RESIDUAL.GENE_EXPRESSION)] = 0
cr = cor(t(RESIDUAL.GENE_EXPRESSION.tmp))
hist(cr, main = 'Distribution of correlation between genes', xlab = 'Correlation')
```

### Clustering residual gene expression
```{r decompse.normalise.data3, fig.height=8, fig.width=8, results='asis'}
# Find principal components of expression to plot
PC <- prcomp(RESIDUAL.GENE_EXPRESSION.tmp, scale.=T, center = T)
